### 2014-11-17 doc/thesis/introduction.tex
Adding text about the typing rules for arithmetic operations

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts evolve into complex programs
\citep{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allows quick development and make it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntaxes and semantics, which usually require a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

@@

Typed Lua's intended use is as an application language, and
we believe that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
checked by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications develop into complex systems rather than just scripts.
We shall see that Typed Lua introduces the refinement of
tables to support the common way that Lua programmers use
them to represent both modules and objects.

We also believe that Typed Lua helps programmers give a more
formal documentation to already existing Lua code, as static types
are also a useful source of documentation in languages that provide
type annotations, because type annotations are always validated by
the type checker and therefore never get outdated.
Thus, programmers can use Typed Lua to define axioms about the
interfaces and types of dynamically typed modules.
We enforce this point by using Typed Lua to statically type
the interface of the Lua standard library and other commonly used
Lua libraries, so our compiler can check Typed Lua code that uses
these libraries.

>>>>

Thus, early error detection and better program structure are two
advantages of static type checking that can lead programmers to
migrate their code from a dynamically typed to a statically
typed language, when their simple scripts evolve into complex programs
\citep{tobin-hochstadt2006ims}.
Dynamically typed languages certainly help programmers during the
beginning of a project, because their simplicity and flexibility
allows quick development and makes it easier to change code according to
changing requirements.
However, programmers tend to migrate from dynamically typed to
statically typed code as soon as the project has consolidated its
requirements, because the robustness of static types helps
programmers link requirements to abstractions.
This migration usually involves different languages that have
different syntaxes and semantics, which usually requires a complete
rewrite of existing programs instead of incremental evolution from
dynamic to static types.

@@

Typed Lua's intended use is as an application language, and
we believe that policies for organizing a program in modules and writing
object-oriented programs should be part of the language and
checked by its optional type system.
An application language is a programming language that helps
programmers develop applications from scratch until these
applications evolve into complex systems rather than just scripts.
We shall see that Typed Lua introduces the refinement of
tables to support the common way that Lua programmers use
them to represent both modules and objects.

We also believe that Typed Lua helps programmers give more
formal documentation to already existing Lua code, as static types
are also a useful source of documentation in languages that provide
type annotations, because type annotations are always validated by
the type checker and therefore never get outdated.
Thus, programmers can use Typed Lua to define axioms about the
interfaces and types of dynamically typed modules.
We enforce this point by using Typed Lua to statically type
the interface of the Lua standard library and other commonly used
Lua libraries, so our compiler can check Typed Lua code that uses
these libraries.

### 2014-11-17 doc/thesis/system.tex
Adding text about the typing rules for arithmetic operations

In this section we will use a reduced core of Typed Lua to present
its typing rules in an easier way.
It is a reduced core of the Lua language with explicit type annotations,
without syntactic sugar, with only if and while control flow statements,
and making variable scope explicit.

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el  \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
a\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} \; | \;
{...} \; | \;
f \; | \;
\{ \; cl \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
a \; | \;
l \; | \;
{<}t{>} \; n\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}]\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
cl ::= & \;\; \mathbf{nothing} \; | \;
\vec{c} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{c}, \vec{e} \; | \;
\vec{c}, me \; | \;
\vec{c}, \vec{e}, me \; | \;
\vec{e}, me\\
me ::= & \;\; a \; | \;
{...}\\
a ::= & \;\; e_{1}(\vec{e_{2}}) \; | \;
e_{1}{:}n(\vec{e_{2}})\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
\mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el \; | \;
\mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
c ::= & \;\; [e_{1}] = e_{2}\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
constructor fields, $me$ are multiple expressions, $a$ are function
applications, $f$ are function declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

@@

>>>>

In this section we use a reduced core of Typed Lua to present the
most interesting typing rules of our type system.
It is also a reduced core of the Lua language with only if and while
control flow statements, without syntactic sugar, with explicit
type annotations, with explicit scope of local variables, and
with explicit declaration of methods.
We use this reduced core because it simplifies the presentation
of our type system.
Appendix \ref{app:rules} presents the full set of typing rules.

\begin{figure}[!ht]
\textbf{Abstract Syntax}\\
\dstart
$$
\begin{array}{rl}
s ::= & \;\; \mathbf{skip} \; | \;
s_{1} \; ; \; s_{2} \; | \;
\vec{l} = el \; | \;
m \; | \;
\mathbf{while} \; e \; \mathbf{do} \; s \; | \;
\mathbf{if} \; e \; \mathbf{then} \; s_{1} \; \mathbf{else} \; s_{2}\\
& | \; \mathbf{local} \; \vec{n{:}t} = el \; \mathbf{in} \; s \; | \;
\mathbf{rec} \; n{:}t = f \; \mathbf{in} \; s \; | \;
\mathbf{return} \; el \; | \;
e(el)_{s} \; | \;
e{:}n(el)_{s}\\
e ::= & \;\; \mathbf{nil} \; | \;
\mathbf{false} \; | \;
\mathbf{true} \; | \;
{\it int} \; | \;
{\it float} \; | \;
{\it string} \; | \;
{...}_{e} \; | \;
f \; | \;
\{ \} \; | \;
\{ \; cl \; \} \\
& | \; e_{1} + e_{2} \; | \;
e_{1} \; {..} \; e_{2} \; | \;
e_{1} == e_{2} \; | \;
e_{1} < e_{2} \; | \;
e_{1} \; \mathbf{and} \; e_{2} \; | \;
e_{1} \; \mathbf{or} \; e_{2} \\
& | \; \mathbf{not} \; e \; | \;
- e \; | \;
\# \; e \; | \;
e(el)_{e} \; | \;
e{:}n(el)_{e} \; | \;
n \; | \;
e_{1}[e_{2}] \; | \;
{<}t{>} \; n\\
l ::= & \;\; n \; | \;
e_{1}[e_{2}] \; | \;
n[e] \; {<}t{>}\\
el ::= & \;\; \mathbf{nothing} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{e}, me\\
cl ::= & \;\; \vec{c} \; | \;
\vec{e} \; | \;
me \; | \;
\vec{c}, \vec{e} \; | \;
\vec{c}, me \; | \;
\vec{c}, \vec{e}, me \; | \;
\vec{e}, me\\
me ::= & \;\; e(el)_{m} \; | \;
e{:}n(el)_{m} \; | \;
{...}_{m}\\
m ::= & \;\; \mathbf{fun} \; n_{1}{:}n_{2} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; n_{1}{:}n_{2} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
f ::= & \;\; \mathbf{fun} \; (){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; ({...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t}){:}r \; s \; ; \; \mathbf{return} \; el\\
& | \; \mathbf{fun} \; (\vec{n{:}t},{...}{:}t){:}r \; s \; ; \; \mathbf{return} \; el\\
c ::= & \;\; [e_{1}] = e_{2}\\
n ::= & \;\; {\it name}\\
\end{array}
$$
\dend
\caption{Typed Lua abstract syntax}
\label{fig:syntax}
\end{figure}

Figure \ref{fig:syntax} presents the abstract syntax of core Typed Lua.
It splits the syntactic categories as follows:
$s$ are statements, $e$ are expressions, $l$ are left-hand values,
$el$ are expression lists, $c$ are constructor fields, $cl$ are lists of
constructor fields, $me$ are expressions with multiple results,
$f$ are function declarations, $m$ are method declarations, $n$ are variable names,
$t$ are first-level types, and $r$ are second-level types.
The notation $\vec{n{:}t}$ denotes the non-empty list
$n_{1}{:}t_{1}, ..., n_{n}{:}t_{n}$.

Our reduced code also splits function application, method application, and
the vararg expression (${...}$) into different categories.
It uses $e(el)_{s}$ to denote function applications that return no value
because they appear as statements,
$e(el)_{e}$ to denote function applications that return only one value,
and $e(el)_{m}$ to denote function applications that return multiple values.
It uses the same categories for method applications, but only ${...}_{e}$
and ${...}_{m}$, as the vararg expression cannot appear as a statement.

@@

Lua has operator overloading, and allows the programmers to redefine
the behavior of some operations.
For instance, programmers can use metatables to redefine the
behavior of arithmetic operations.
Even though Typed Lua does not support operator overloading yet,
it includes typing rules that allow programmers to use the
dynamic type when they are using overloaded operations.
The following typing rules show how Typed Lua uses the dynamic type
to handle the overloading of arithmetic operations:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-ARITH5}\\
\dfrac{\env_{1} \vdash e_{1}:\Any, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3}}
      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
\end{array}
\;
\begin{array}{c}
\mylabel{T-ARITH6}\\
\dfrac{\env_{1} \vdash e_{1}:t, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:\Any, \env_{3}}
      {\env_{1} \vdash e_{1} + e_{2}:\Any, \env_{3}}
\end{array}
\end{array}
\]

The rule \textsc{T-ARITH5} allows type checking the following
example:
\begin{center}
\begin{tabular}{l}
$\mathbf{local} \; x{:}\Any = 1 \; \mathbf{in} \; x = x + 1$
\end{tabular}
\end{center}

This example is safe, but the following it is not:
\begin{center}
\begin{tabular}{l}
$\mathbf{local} \; x{:}\Integer, \;y{:}\Any = 1 \; \mathbf{in} \; x = x + y$
\end{tabular}
\end{center}

\textit{/* Expression list rules. Where should we describe the
discarding/replacement of tuple elements? */}

Even though Lua has multiple assignments, the rule \textsc{T-ASSIGNMENT}
shows that tuples types and consistent-subtyping make straightforward
to type check them.
\[
\begin{array}{c}
\mylabel{T-ASSIGNMENT}\\
\begin{array}{c}
\dfrac{\env_{1} \vdash el:r_{1}, \env_{2} \;\;\;
       \env_{2} \vdash \vec{l}:r_{2}, \env_{3} \;\;\;
       r_{1} \lesssim r_{2}}
      {\env_{1} \vdash \vec{l} = el:\env_{3}}
\end{array}
\end{array}
\]

Multiple assignments and multiple return values can also appear in
function applications, as the following example shows:
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; m:\Void \rightarrow \Integer \times \String =$} \\
& \multicolumn{3}{l}{$\mathbf{fun} \; ():\Integer \times \String$} \\
& & \multicolumn{2}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; 2, ``foo"$} \\
\multicolumn{4}{l}{$\mathbf{in}$} \\
& \multicolumn{3}{l}{$\mathbf{local} \; s:\Integer \times \Integer \rightarrow \Integer =$} \\
& & \multicolumn{2}{l}{$\mathbf{fun} \; (x:\Integer, y:\Integer):\Integer$} \\
& & & \multicolumn{1}{l}{$\mathbf{skip} \;;\; \mathbf{return} \; x + y$} \\
& \multicolumn{3}{l}{$\mathbf{in} \; s(m()_{e}, m()_{m})_{s}$}
\end{tabular}
\end{center}

The $\mathbf{or}$ idiom.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; x:\String \cup \Nil = \mathbf{nothing} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; y:\String = x \; \mathbf{or} \; ``Hello"$}
\end{tabular}
\end{center}

Overloading on the input parameter.
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; o:\String \times \String \cup \Integer \rightarrow \String =$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; (a:\String, b:\String \cup \Integer):\String$}\\
& & \multicolumn{2}{l}{$\mathbf{if} \; type(b) == ``string" \; \mathbf{then}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; a \; {..} \; b$}\\
& & \multicolumn{2}{l}{$\mathbf{else}$}\\
& & & \multicolumn{1}{l}{$\mathbf{return} \; rep(a, b)_{e}$}\\
\multicolumn{4}{l}{$\mathbf{in} \; o(``foo", 2)_{s}$}
\end{tabular}
\end{center}

We are assuming that the function \emph{rep} is in the environment and has type
$\String \times \Integer \times \String \cup \Nil \rightarrow \String$.

Overloading on the return type.
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; q:\pi_{1}, r:\pi_{2} = idiv(1, 2) \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{if} \; type(r) == ``string" \; \mathbf{then}$}\\
& & \multicolumn{1}{l}{$print(``ERROR: " \; .. \; r)_{s}$}\\
& \multicolumn{2}{l}{$\mathbf{else}$}\\
& & \multicolumn{1}{l}{$print(q + r)_{s}$}
\end{tabular}
\end{center}

We are assuming that functions \emph{idiv} and \emph{print} are in
the environment with the respective types
$\Integer \times \Integer \rightarrow \Integer \times \Integer \sqcup \Nil \times \String$
and
$\Value{*} \rightarrow \Void$.

Table type as hash.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\String:\Integer \cup \Nil\}_{closed} = \{ [``foo"] = 1 \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; x:\Integer \cup \Nil = t[``foo"] \; \mathbf{in} \; x = t[``bar"]$}
\end{tabular}
\end{center}

Table type as array.
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\Integer:\String \cup \Nil\}_{closed} = \{ ``a", ``b", ``c" \} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$\mathbf{local} \; x:\String \cup \Nil = t[1] \; \mathbf{in} \; x = t[4]$}
\end{tabular}
\end{center}

Table type as record (first example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``l"] = ``Reed", [``f"] = ``Lou" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Table type as record (second example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``f"] = ``Lewis", [``m"] = ``Allan", [``l"] = ``Reed" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Table type as record (third example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{``f":\String, ``m":\String \cup \Nil, ``l":\String\}_{closed} =$}\\
& \multicolumn{1}{l}{$\{ [``f"] = ``Lou", [``l"] = ``Reed" \}$}\\
\multicolumn{2}{l}{$\mathbf{in} \; \mathbf{skip}$}
\end{tabular}
\end{center}

Refinement of table types (first example).
\begin{center}
\begin{tabular}{ll}
\multicolumn{2}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
& \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
\end{tabular}
\end{center}

Refinement of table types (second example).
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; t:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = p \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$t[``f"] \; {<}\String{>} = ``Lou";$}\\
& & \multicolumn{1}{l}{$t[``l"] \; {<}\String{>} = ``Reed";$}\\
\end{tabular}
\end{center}

Objects.
\begin{center}
\begin{tabular}{llll}
\multicolumn{4}{l}{$\mathbf{local} \; s:\{``x:"\Number, ``y":\Number\}_{open} = \{ [``x"] = 0.0, [``y"] = 0.0 \}$}\\
\multicolumn{4}{l}{$\mathbf{in}$}\\
& \multicolumn{3}{l}{$\mathbf{fun} \; s{:}new (x:number, y:number):\Self$}\\
& & \multicolumn{2}{l}{$\mathbf{local} \; s:\{``x":\Number, ``y":\Number\}_{closed} =$}\\
& & & \multicolumn{1}{l}{$setmetatable(\{\}, \{ [``\string_\string_index"] = self \}$}\\
& & \multicolumn{2}{l}{$\mathbf{in} \; s[``x"] = x; \; s[``y"] = y; \; \mathbf{return} \; s$}\\
; & \multicolumn{3}{l}{$\mathbf{fun} \; s{:}move (x:number, y:number):\Void$}\\
& & \multicolumn{2}{l}{$self[``x"] = self[``x"] + x;$}\\
& & \multicolumn{2}{l}{$self[``y"] = self[``y"] + y$}
\end{tabular}
\end{center}
